Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/overview.tex
tightened up 2.0

We present an overview of our approach by considering an example.  To
highlight the ideas, we abstract over some technical details, which
are fully formalized in the rest of the paper.  The example itself is
also implemented in our implementation.

\paragraph{Computing order statistics.}

@@

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected $O(n)$ time.
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate: we wish to be able to map \texttt{qs} to
over another list and do so efficiently.

One way to regain efficiency is to use a multi-pass algorithm that
performs frequently performed computations first so that they can be
reused.  It is nontrivial what such computations are in this example
but after some reflection, we can realize that they are the
comparisons performed by the partition function.  To minimize them, we
can pre-sort the input list \texttt{l} into a sorted list \texttt{s}
in the first pass, and then perform lookups in \texttt{s} in the
second pass.  Unfortunately, since \texttt{s} is a list, a lookup
would require linear time (also on average), leading to no improvement
in efficiency.  But this is not too hard: after sorting \texttt{s} we
can copy in into an array and use binary search to find the element
with the desired rank.  As another option, we can construct a binary
search tree from the input and perform size-based searches on the
binary search tree to find the elements with the desired rank.  Such
an approach would fit nicely into the functional paradigm but requires
one more additional algorithmic ingenuity: to ensure efficiency we
would need to store at each note the size of its subtree (these are
sometimes called ``augmented trees'').  In summary, we are able to
improve efficiency by replacing our simple solution with a multi-pass
algorithm that first preprocesses the input to generate a lookup data
structure and then in the second pass performs fast lookups.  As
outlined above, in addition, this transformation is highly non-trivial
as it involves reasoning about intricate algorithmic concerns, and
implementing more complex algorithms and data structures.  Both of
these approaches however improve the run-time to (expected)
$\Theta(n\log{n} + m\log{n})$, a near linear-time improvement in $m$.

Such transformations, called pass-separation by J{\o}rring and
Scherlis are commonly employed.  For example, as briefly mentioned in
\ref{sec:intro}, modern graphics software is written exactly in this
way.  Unfortunately, as the example illustrates, multipass programs
are naturally complex.  We thus ask: is it possible to
perform such transformations automatically.  We answer this question
affirmatively.  Our approach is to start with a staged language and
then use a splitting algorithm to generate multi-pass programs from
staged programs.

@@

>>>>

@@

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  Assuming that the input is uniformly
randomly ordered (which can be achieved by prepermuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

\pagebreak
We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{partition} depend only on \texttt{l}, although different calls to
\texttt{partition} are made by each call to \texttt{qs}.

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---storing at each node the size of its left subtree (i.e., using
\emph{augmented trees}). This approach has (expected) runtime $\Theta(n\log{n} +
m\log{n})$, a nearly linear-time improvement in $m$.

In both cases, we replace our simple solution with a two-pass algorithm which
first processes the input into a lookup data structure, and then subsequently
performs fast lookups. Improving algorithms in this way is highly non-trivial,
as it requires reasoning about intricate algorithmic concerns, and implementing
more complex algorithms and data structures.

In this paper, we present a splitting algorithm which transforms staged programs
into multi-pass programs. In certain cases, this algorithm \emph{automatically}
improves algorithms in the way described above.

%Such transformations, called pass-separation by J{\o}rring and
%Scherlis are commonly employed.  For example, as briefly mentioned in
%\ref{sec:intro}, modern graphics software is written exactly in this
%way.  Unfortunately, as the example illustrates, multipass programs
%are naturally complex.

@@

%The example itself is also implemented in our implementation.
